---
title: "Covid-19 and Economics"
subtitle: "... where are the links?"
author: "Ralf Becker"
date: "June 2020"
output: 
  beamer_presentation:
    includes:
      in_header: ../latex_template.tex
#     in_header: ../latex_student.tex  # use for student version
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE,  comment=NA)
```

```{r}
library(sets)         # used for some set operations
library(forecast)     # used for some data smoothing
library(readxl)       # enable the read_excel function
library(tidyverse)    # for almost all data handling tasks
library(ggplot2)      # plotting toolbox
library(xts)
```

## Aim for today

\begin{itemize}
  \item Point out why this is relevant for Economists
  \item Show you some of the up to date Covid-19 data
  \item Show you some of the data analysis skills you will learn
  \item Show you some of the data coding skills you will learn
\end{itemize}

What you see today is related to a larger data project in the \href{https://www.core-econ.org/}{CORE-ECON Covid-19 Collection}.

## Why, as economists, should we look at Covid-19

The **small** questions

* Understanding the current and future needs are important for business and government for planning (toilet paper producers, fresh food importers, pasta retailers, NHS hospitals, etc) 

The **big** questions

Covid-19 highlights other important issues we ought to think about as economists: inequality, climate change, ...

* What is the value of a life?
* Why are some groups in society affected more than others?
* Rebalancing between Markets, Government, Civil Society


## Is there a society?

\begin{columns}
  
  \begin{column}{0.5\textwidth}
  "[...] there's no such thing as society. There are individual men and women and there are families. And no government can do anything except through people, and people must look after themselves first. It is our duty to look after ourselves and then, also, to look after our neighbours." (Margaret Thatcher, 1987)

  \end{column}
  \begin{column}{0.5\textwidth}
    \begin{figure}
    	\centering
    	\includegraphics[width=5cm]{Thatcher.jpg}\\
    \end{figure}

  \end{column}
    
\end{columns}

 

Often the world is portrayed as \textcolor{blue}{markets} on one hand and \textcolor{red}{government} on the other. 

## \textcolor{blue}{Markets},\textcolor{red}{Government} and \textcolor{green}{Society}

For long economists have realised that there is more than the simple juxtaposition of optimising self-interested individuals against interventionist governments.



\begin{columns}
  
  \begin{column}{0.5\textwidth}
    \begin{figure}
    	\centering
    	\includegraphics[width=3cm]{Ostrom.jpg}\\
    \end{figure}
    Elinor Ostrom\\ 
    Economics of the Commons
  \end{column}
  \begin{column}{0.5\textwidth}
    \begin{figure}
    	\centering
    	\includegraphics[width=3cm]{Kahneman.jpg}\\
    \end{figure}
    Daniel Kahnemann\\
    Behavioural Economics
  \end{column}
    
\end{columns}


```{r, Import Mobility Data}
# Data from 
# https://www.google.com/covid19/mobility/
mob_data <- read.csv("Global_Mobility_Report.csv")
mob_data <- mob_data %>% filter(sub_region_1 %in%       c("Greater Manchester",                             "Stockholm County", "Berlin"))
mob_data$date <-        as.Date(as.character(mob_data$date),"%Y-%m-%d")
```

## Covid-19 and Society

The current crisis delivers some excellent examples of how markets, governments and society work together. 

\footnotesize

```{r, Plot Mobility, fig.height = 3, fig.width = 5}
ggplot(mob_data,aes(x =date, y=transit_stations_percent_change_from_baseline, color=sub_region_1)) + geom_line(size = 1.2) +
  labs(title = "Comparative use Public Transport", caption = "Source: https://www.google.com/covid19/mobility/") +
  ylab("Change compared to same week previous year") +     
  scale_color_discrete(name="Region")
```

\normalsize



```{r, Import Stringency Data}
# Data from https://www.bsg.ox.ac.uk/research/research-projects/coronavirus-government-response-tracker
policy_data <- read.csv("https://raw.githubusercontent.com/OxCGRT/covid-policy-tracker/master/data/OxCGRT_latest.csv")
policy_data <- policy_data %>% filter(CountryName %in% c("United Kingdom", "Sweden", "Germany"))
policy_data$Date <- as.Date(as.character(policy_data$Date),"%Y%m%d")
policy_data <- droplevels(policy_data)  # remove unused levels
policy_data$CountryName <- factor(policy_data$CountryName, levels = c("Germany","United Kingdom","Sweden"))
```

## Covid-19 and Society

So Germany and Sweden have fairly similar responses (wrt to the use of public transport). Is that because they had similar policies?

\footnotesize

```{r, Plot Policy, fig.height = 3, fig.width = 5}
ggplot(policy_data,aes(x =Date, y=StringencyIndex, color=CountryName)) + 
  geom_line(size = 1.2) +
  labs(title = "Stringency of preventive policy measures", caption = "Source: Univ of Oxford, Blavatnik Schoool of Government") +
  ylab("Stringency Index") + 
  scale_color_discrete(name="Country")
```

\normalsize

## Policy and Behaviour 

This is an example where the policy alone does not explain the differences in behaviour.

Despite much softer restrictions (compared to Germany) Swedes have restricted their use of public transport in a similar manner to their German counterparts. 

Human behaviour is not easliy modeled!

How messages are communicated matters. The UK and other governments now employ Behavioural Economists and Psychologists to advise them. 

See the Behavioural Insights Team (https://www.bi.team/our-work/covid-19/).


## Again, why is this economics?

Typically, after huge crisis, we learn, and we change.

* The great depresison: Governments had a role in aggregate demand management
* WWII: The Bretton Woods Institutions were created in order to co-operate, and the Marshall Plan helped Germany
* Financial Crisis, 2008: Reconsideration of the role of financial markets and regulation

What economic changes will come after Covid-19?

Is it another international institution to deal with global risks (pandemics, climate change, etc.)?

How should we decide which of these potentially catastrophic events we should tackle, if any, or both?

## Can we apply simple economic principles to catastrophes?

There are certainly elements of traditional cost-benefit analysis that could be applied, but ...  

&nbsp;

\begin{columns}
  
  \begin{column}{0.33\textwidth}
    mostly measureable in monetary terms
    \begin{enumerate}
      \item ventilators
      \item PPE
      \item labs
      \item scientists to develop vaccine
      \item bureaucracies to coordinate. etc
    \end{enumerate}
  \end{column}
  \begin{column}{0.33\textwidth}
     \begin{figure}
        	\centering
        	\includegraphics[width=4cm]{scale_CB.jpg}\\
      \end{figure}

  \end{column}
  \begin{column}{0.33\textwidth}
    many which are not easily measured in monetary terms
        \begin{enumerate}
      \item saved lifes, now and in the future
    \end{enumerate}

  \end{column}
    
\end{columns}

\tiny

Literature:

* Martin and Pindyck. "Averting catastrophes: The strange economics of Scylla and Charybdis." American Economic Review 105, no. 10 (2015): 2947-85.
* Viscusi and Aldy, “The value of a statistical life: a critical review of market estimates throughout the world,” Journal of risk and uncertainty, 2003, 27 (1), 5–76.

\normalsize

## How will your degree help you think?

* We will not give you complete answers ... where there are none
* We will give you some tools to structure your thinking process

\begin{figure}
\begin{tabular}{cc}
  \includegraphics[width=30mm]{maths} &   \includegraphics[width=30mm]{models} \\
 \includegraphics[width=30mm]{coding} &   \includegraphics[width=30mm]{communicating} 
\end{tabular}
\end{figure}

##

Thanks for joining us!

any questions? don't hold back!
