---
title: "Demonstration Class 1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```


# Preparing your workfile

We add the basic libraries needed for this week's work: 

```{r}
library(tidyverse)    # for almost all data handling tasks
library(readxl)       # to import Excel data
library(ggplot2)      # to produce nice graphiscs
library(stargazer)    # to produce nice results tables
library(haven)        # to import stata file
library(AER)          # access to HS robust standard errors
library(estimatr)     # use robust se
source("stargazer_HC.r")
```


# Introduction

The data are an extract from the [Understanding Society Survey](https://www.understandingsociety.ac.uk/) (formerly the British Household Survey Panel).

# Data Upload - and understanding data structure

Upload the data, which are saved in a STATA datafile (extension `.dta`). There is a function which loads STATA file. It is called `read_dta` and is supplied by the `haven` package.

```{r}
data_USoc <- read_dta("20222_USoc_extract.dta")
data_USoc <- as.data.frame(data_USoc)    # ensure data frame structure
names(data_USoc)
```

Let us ensure that categorical variables are stored as `factor` variables. It is easiest to work with these in R.

```{r}
data_USoc$region <- as_factor(data_USoc$region)
data_USoc$male <- as_factor(data_USoc$male)
data_USoc$degree <- as_factor(data_USoc$degree)
data_USoc$race <- as_factor(data_USoc$race)
```

The pay information (`paygu`) is provided as a measure of the (usual) gross pay per month. As workers work for dy we shall also adjust for increasing price levels ( as measured`mutate` function. We call this variable `hrpay` and also calculate the natural log of this variable (`lnhrpay`).

```{r}
data_USoc <- data_USoc %>% 
              mutate(hrpay = paygu/(jbhrs*4)/(cpi/100)) %>% 
              mutate(lnhrpay = log(hrpay))
```


We will also calculate a modified race variable

```{r}
data_USoc %>% count(race)
```

We shall create a variable called `white` which re-groups this into two categories, white and non-white.

```{r}
data_USoc <- data_USoc %>%
    mutate(white = fct_recode(race,
    "white"    = "white",    # new level = old level
    "non-white"  = "mixed",
    "non-white"  = "asian",
    "non-white"  = "black",
    "non-white"  = "other"))
```

Let's estimate a regression, allowing average hourly pay to differ between whites and non-whites

```{r}
mod1 <- lm(lnhrpay~white,data = data_USoc)
stargazer_HC(mod1,type="text")
```


Now we calculate inflation adjusted figures for gross monthly pay (`paygu`) and scale it to annual pay (`annualpay`).

```{r}
data_USoc <- data_USoc %>% 
              mutate(annualpay = paygu*12/(cpi/100))
summary(data_USoc$annualpay)
```

Now we look at estimating the following model

\[anualpay = \beta_0 + \beta_1 ~ age + v\]

```{r}
mod1 <- lm(annualpay~age,data = data_USoc,subset=(degree == "first degree"))
mod2 <- lm(annualpay~age,data = data_USoc,subset=(degree == "no degree"))
stargazer(mod1,mod2, type = "text", 
          column.labels = c("first degree","without degree"),omit.stat = "all") 
```

Now we use these models to predict annual wages at the age of 20 and 65 depending on whether someone has a degree or not.

```{r}
# with degree, use mod1
y20_1 <- mod1$coefficients[1]+mod1$coefficients[2]*20
y65_1 <- mod1$coefficients[1]+mod1$coefficients[2]*65

# without degree, use mod2
y20_0 <- mod2$coefficients[1]+mod2$coefficients[2]*20
y65_0 <- mod2$coefficients[1]+mod2$coefficients[2]*65
```

Let's create a plot with the two fitted regression lines. But before we do that we create a temporary dataset which only contaisn obs with the two `degree` outcomes we consider here, i.e. "no degree" and "first degree".

```{r}
temp <- data_USoc %>% filter(degree %in% c("no degree", "first degree"))
ggplot(temp, aes(x=age,y=annualpay,color = degree))+
  geom_point(size=0.01) +
  geom_smooth(method = "lm", se = FALSE)

```

To calculate the lifetime earnings premium for a graduate (earnings between 20 and 65) we calculate the lifetime earnings of a graduate and substract the lifetime earnings of a non-graduate.

```{r}
L1 = (65-20)*(y20_1+y65_1)/2
L0 = (65-20)*(y20_0+y65_0)/2
D = L1 - L0
unname(L1)   # If you don't use unname then
unname(L0)   # all these values will inherit the
unname(D)    # name "intercept" from coefficients

```

So this almost half a million pounds over a lifetime. Is that the value of a degree?