---
title: "Introduction to Regression Analysis 1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```


# Preparing your workfile

We add the basic libraries needed for this week's work: 

```{r}
library(tidyverse)    # for almost all data handling tasks
library(readxl)       # to import Excel data
library(ggplot2)      # to produce nice graphiscs
library(stargazer)    # to produce nice results tables
library(haven)        # to import stata file
library(AER)          # access to HS robust standard errors
library(estimatr)     # use robust se
```


# Introduction



# Data Upload - and understanding data structure

Upload the data, which are saved in a STATA datafile (extension `.dta`). There is a function which loads STATA file. It is called `read_dta` and is supplied by the `haven` package.

```{r}
data_USoc <- read_dta("20222_USoc_extract.dta")
data_USoc <- as.data.frame(data_USoc)    # ensure data frame structure
names(data_USoc)
```

Let us ensure that categorical variables are stored as `factor` variables. It is easiest to work with these in R.

```{r}
data_USoc$region <- as_factor(data_USoc$region)
data_USoc$male <- as_factor(data_USoc$male)
data_USoc$degree <- as_factor(data_USoc$degree)
data_USoc$race <- as_factor(data_USoc$race)
```

Calculate log houry pay

```{r}
data_USoc <- data_USoc %>% 
              mutate(hrpay = paygu/(jbhrs*4)/(cpi/100)) %>% 
              mutate(lnhrpay = log(hrpay))
```


# A regression - Gender differences

Testing whether we can add robust errors to stargazer output:

```{r}
mod1 <- lm(lnhrpay~male,data = data_USoc)
cov1 <- vcovHC(mod1, type = "HC1")
robust_se <- sqrt(diag(cov1))
stargazer(mod1,type="text",se=list(NULL, robust_se))
```

That works well. But I sort of want to avoid students having to write all these lines if they want to use robust ses ... after all we want to make that standard. So I decided to write a little function

```{r}
stargazer_HC <- function(mod, type_in = "text") {
  cov1 <- vcovHC(mod, type = "HC1")
  robust_se <- sqrt(diag(cov1))
  stargazer(mod,type=type_in,se=list(NULL, robust_se))
  # I actually don't fully understand why we need to add the NULL list, but it doesn't work without
}
```

And that works well. All students need to call is.

```{r}
mod1 <- lm(lnhrpay~male,data = data_USoc)
stargazer_HC(mod1)
```

One of the great features of stargazer is that you can plot several regression results next to each other.


```{r}
mod2 <- lm(lnhrpay~educ,data = data_USoc)
mod3 <- lm(lnhrpay~male+educ,data = data_USoc)
stargazer(mod1,mod2,mod3, type = "text")
```

Alternatively one can hand in a list of models

```{r}
stargazer(list(mod1,mod2,mod3), type = "text")
```

Let's create the robust standard errors for the two additional models

```{r}
cov2 <- vcovHC(mod2, type = "HC1")
robust_se2 <- sqrt(diag(cov2))
cov3 <- vcovHC(mod3, type = "HC1")
robust_se3 <- sqrt(diag(cov3))
```

Now we can feed this int the stargazer function

```{r}
stargazer(list(mod1,mod2), se=list(robust_se,robust_se2), type ="text")
```


I am now trying to produce a version of `stargazer_HC(mod1)` which allows for multiple models and robust errors in these, but so far I am failing. Here is my best attempt:

```{r}
stargazer_HCm <- function(mod_all, type_in = "text") {
# model collection should come in as a list of models
  n_mod <- length(mod_all)
  rep_se <- list(NULL)  # use this to replace se
  i <- 1        # count the number of models 
  
  if (length(mod_all) == 1) {
    cov1 <- vcovHC(mod_all[[1]], type = "HC1") # calculate next rob se
    robust_se <- sqrt(diag(cov1))
    stargazer(mod,type="text",se=list(NULL,robust_se))
  } else {
    for (mod_i in mod_all){
    covi <- vcovHC(mod_i, type = "HC1") # calculate next rob se
    robust_se <- sqrt(diag(covi))
    rep_se[[i]] <- robust_se  # add next robust standard error to rep_se list
    i <- i + 1
    }

    stargazer(mod_all, type = "text", se=rep_se)
  }
}
```

```{r}
stargazer_HCm(list(mod1,mod2,mod3))
```

This works. How do we now manage to make the call to the function `stargazer_HCm(mod1,mod2,mod3)` instead of `stargazer_HCm(list(mod1,mod2,mod3))`, i.e. I want to avoid the list. Let's see whether the `...`function input works

```{r}
test_list <- function(...){
  list(...)
}
test_list(robust_se,robust_se2)
```

So, that has indeed a prduced a list. So let's apply this to our problem

```{r}
stargazer_HCm2 <- function(..., type_out = "text", type_HC = "HC1") {
  # the first inputs are the models which are to be displayed
  # type_out carries through to the stargazer function, default: "text", other; "latex", "html"
  # type_HC carries through tothe vcovHC function, default: "HC1" (or Stata equivalent)
  mod_all <- list(...)  # model collection should be a list of models
  n_mod <- length(mod_all)
  rep_se <- list(NULL)  # use this to replace se
  i <- 1        # count the number of models 
  
  if (length(mod_all) == 1) {
    cov1 <- vcovHC(mod_all[[1]], type = type_HC) # calculate next rob se
    robust_se <- sqrt(diag(cov1))
    stargazer(mod_all[[1]],type=type_out,se=list(NULL,robust_se))
  } else {
    for (mod_i in mod_all){
    covi <- vcovHC(mod_i, type = type_HC) # calculate next rob se
    robust_se <- sqrt(diag(covi))
    rep_se[[i]] <- robust_se  # add next robust standard error to rep_se list
    i <- i + 1
    }

    stargazer(mod_all, type =type_out, se=rep_se)
  }
}
```

```{r}
stargazer_HCm2(mod1,mod2)
```

excellent this works, can 


Interestingly I found a package (`estimatr`) which includes a function to easily run regerssions with robust errors

```{r}
lm_robust(lnhrpay~educ,data = data_USoc, se_type = "HC1")
```

However, the output is not compatible with stargazer.

Any ideas?

