---
title: "Introduction to Regression Analysis 1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```


# Preparing your workfile

We add the basic libraries needed for this week's work: 

```{r}
library(tidyverse)    # for almost all data handling tasks
library(readxl)       # to import Excel data
library(ggplot2)      # to produce nice graphiscs
library(stargazer)    # to produce nice results tables
library(haven)        # to import stata file
library(AER)          # access to HS robust standard errors
library(estimatr)     # use robust se
```


# Introduction



# Data Upload - and understanding data structure

Upload the data, which are saved in a STATA datafile (extension `.dta`). There is a function which loads STATA file. It is called `read_dta` and is supplied by the `haven` package.

```{r}
data_USoc <- read_dta("20222_USoc_extract.dta")
data_USoc <- as.data.frame(data_USoc)    # ensure data frame structure
names(data_USoc)
```

Let us ensure that categorical variables are stored as `factor` variables. It is easiest to work with these in R.

```{r}
data_USoc$region <- as_factor(data_USoc$region)
data_USoc$male <- as_factor(data_USoc$male)
data_USoc$degree <- as_factor(data_USoc$degree)
data_USoc$race <- as_factor(data_USoc$race)
```

Calculate log houry pay

```{r}
data_USoc <- data_USoc %>% 
              mutate(hrpay = paygu/(jbhrs*4)/(cpi/100)) %>% 
              mutate(lnhrpay = log(hrpay))
```


# A regression - Gender differences

Testing whether we can add robust errors to stargazer output:

```{r}
mod1 <- lm(lnhrpay~male,data = data_USoc)
cov1 <- vcovHC(mod1, type = "HC1")
robust_se <- sqrt(diag(cov1))
stargazer(mod1,type="text",se=list(NULL, robust_se))
```

That works well. But I sort of want to avoid students having to write all these lines if they want to use robust ses ... after all we want to make that standard. So I decided to write a little function

```{r}
stargazer_HC <- function(mod, type_in = "text") {
  cov1 <- vcovHC(mod, type = "HC1")
  robust_se <- sqrt(diag(cov1))
  stargazer(mod,type=type_in,se=list(NULL, robust_se))
  # I actually don't fully understand why we need to add the NULL list, but it doesn't work without
}
```

And that works well. All students need to call is.

```{r}
mod1 <- lm(lnhrpay~male,data = data_USoc)
stargazer_HC(mod1)
```

One of the great features of stargazer is that you can plot several regression results next to each other.


```{r}
mod2 <- lm(lnhrpay~educ,data = data_USoc)
mod3 <- lm(lnhrpay~male+educ,data = data_USoc)
stargazer(mod1,mod2,mod3, type = "text")
```

Alternatively one can hand in a list of models

```{r}
stargazer(list(mod1,mod2,mod3), type = "text")
```

I am now trying to produce a version of `stargazer_HC(mod1)` which allows for multiple models and robust errors in these, but so far I am failing. Here is my best attempt:

```{r}
stargazer_HCm <- function(mod, type_in = "text") {
# model collection should come in as a list of models
  rep_se <- list(NULL)  # use this to replace se

  # if extra models process them here
  for (mod_i in mod_all){
    cov1 <- vcovHC(mod_i[[1]], type = "HC1") # calculate next rob se
    robust_se <- sqrt(diag(cov1))
    rep_se <- list(rep_se, list(robust_se))  # add next se
  }

  stargazer(mod,type="text",se=rep_se)
}
```

```{r}
stargazer_HCm(list(mod1,mod2,mod3))
```

This doesn't work. 

Interestingly I found a package (`estimatr`) which includes a function to easily run regerssions with robust errors

```{r}
lm_robust(lnhrpay~educ,data = data_USoc, se_type = "HC1")
```

However, the output is not compatible with stargazer.

Any ideas?