---
title: "Introduction to Regression Analysis - Multivariate Regression"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```


# Preparing your workfile

We add the basic libraries needed for this week's work: 

```{r}
library(tidyverse)    # for almost all data handling tasks
library(ggplot2)      # to produce nice graphiscs
library(stargazer)    # to produce nice results tables
library(haven)        # to import stata file
library(AER)          # access to HS robust standard errors
source("stargazer_HC.r")  # includes the robust regression display
```


# Introduction

The data are an extract from the [Understanding Society Survey](https://www.understandingsociety.ac.uk/) (formerly the British Household Survey Panel).

# Data Upload - and understanding data structure

Upload the data, which are saved in a STATA datafile (extension `.dta`). There is a function which loads STATA file. It is called `read_dta` and is supplied by the `haven` package.

```{r}
data_USoc <- read_dta("20222_USoc_extract.dta")
data_USoc <- as.data.frame(data_USoc)    # ensure data frame structure
names(data_USoc)
```

Let us ensure that categorical variables are stored as `factor` variables. It is easiest to work with these in R.

```{r}
data_USoc$region <- as_factor(data_USoc$region)
data_USoc$male <- as_factor(data_USoc$male)
data_USoc$degree <- as_factor(data_USoc$degree)
data_USoc$race <- as_factor(data_USoc$race)
```

Click on the little table symbol in your environment tab to see the actual data table. 



The pay information (`paygu`) is provided as a measure of the (usual) gross pay per month. As workers work for dy we shall also adjust for increasing price levels ( as measured`mutate` function. We call this variable `hrpay` and also calculate the natural log of this variable (`lnhrpay`).

```{r}
data_USoc <- data_USoc %>% 
              mutate(hrpay = paygu/(jbhrs*4)/(cpi/100)) %>% 
              mutate(lnhrpay = log(hrpay))
```

As we wanted to save these additional variables we assign the result of the operation to `data_USoc`.

Let's run a simple regression of `lnhrpay` on `educ` which is a variable which counts the years of formal education.

```{r}
mod1 <- lm(lnhrpay~educ,data = data_USoc)
stargazer_HC(mod1)
```

Before we continue we create a squared age variable

```{r}
data_USoc <- data_USoc %>% mutate(agesq = age*age/100)
```

Now we allow for age as an additional explanatory variable. But we will allow for the effect of `age` to be nonlinear, in particular quadratic.

```{r}
mod2 <- lm(lnhrpay~educ+age+agesq,data = data_USoc)
stargazer_HC(mod1,mod2)
```


