---
title: "Introduction to Data Handling and Statistics"
author: "Ralf Becker"
date: "29 November 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Preparing your workfile

R is a powerful software for statistical analysis. It is open source and hence FREE software. It is constantly developped and functionality is being improved. The "price" we pay for this is that we have to do a little more set-up work to make it work.

In particular we need packages which are provided for free by researchers/programmers that provide useful functionality.

But we add libraries which enhance its capabilities. 

```{r}
library(tidyverse)    # for almost all data handling tasks
library(readxl)       # to import Excel data
library(ggplot2)      # to produce nice graphiscs
library(stargazer)    # to produce nice results tables
```


# Introduction

Here we use an example which, at the time, made a lot of waves.  Carmen Reinhart and Kenneth Rogoff (2010) wrote work that attempted to examine the [relationship between Debt to GDP levels and GDP growth](https://scholar.harvard.edu/files/rogoff/files/growth_in_time_debt_aer.pdf). Their general conclusion was that, as long as the debt to GDP ration does not exceed 90% there seems to be no clear relationship between the two. However, countries eunning higher debt to GDP ratios are paying a sizeable growth penalty.

The empirical work of Reinhard and Rogoff was critised on three aspects by Thomas Herndon, Michael Ash and Robert Pollin. This work was published in the [Cambridge Journal of Economics in 2014](https://academic.oup.com/cje/article-abstract/38/2/257/1714018?redirectedFrom=fulltext) and previous to that as a working paper. The [working paper's website](https://www.peri.umass.edu/publication/item/526-does-high-public-debt-consistently-stifle-economic-growth-a-critique-of-reinhart-and-rogoff) also contains all the relevant data and code.

# Importing Data

This is a simplified spreadsheet (based on the work of Herndon et al.)

```{r}
RRData <- read_excel("RRdata.xlsx")
str(RRData)  # prints some basic info on variables
```

We can see that three variables are numeric (`num`) variables which is as we expect. The fourth variable `Country` is labeled as a character (`chr`) variable, basically text. It will pay off for later to indicate to R that this is a categorical (nominal) variable. This is done as follows:

```{r}
RRData$Country <- as.factor(RRData$Country)
str(RRData)
```

There are 1171 country-year observations. In the dataset are data from 20 countries:

```{r}
unique(RRData$Country)
```

And data from 1946 to 2009.

```{r}
summary(RRData$Year)
```

`debtgdp` is the debt to GDP ratio

```{r}
summary(RRData$debtgdp)
```

`dRGDP` is GDP growth rate

```{r}
summary(RRData$dRGDP)
```


# An initial basic data plots

Let's have a look at some of the data. We will first look at a scatterplot (using the amazing `ggplot` library).

```{r}
p1 <- ggplot(RRData,aes(debtgdp,dRGDP)) +
  geom_point() +    # this produces the scatter plot
  geom_smooth(method = "lm", se = FALSE)  # this adds the linear line of best fit
p1
```

We could actually run this regression

```{r}
mod1 <- lm(dRGDP~debtgdp, data= RRData)
summary(mod1)   # this is the standard R way for results
stargazer(mod1,type="text") # a nicer way
```

There are issues with running a regression like this. Perhaps most importantly that the observations cannot be argued to be independent. The growth rate and debt to GDP ratios of, say, the UK in 1981 is clearly not independent of the data in 1980. In particular debt to GDP ratios are slow moving data.

To illustrate this point we should look at the data as time series.

```{r}
tempdata <- RRData %>% filter(Country %in% c("Germany","Greece","UK","US"))
ggplot(tempdata,aes(Year,debtgdp,color=Country)) +
  geom_line()     # this produces the line plot

tempdata <- RRData %>% filter(Country %in% c("Germany","Greece","UK","US"))
ggplot(tempdata,aes(Year,dRGDP,color=Country)) +
  geom_line(size=1)     # this produces the line plot

```

You can clearly see that the `debtgdp` data, from year to year, are dependent. Also, the `dRGDP` plot reveals that there is a fair bit of correlation between the growth rates in economies.

# Some data cuts

Let's look at the data for a few countries

```{r}
tempdata <- RRData %>% filter(Country %in% c("Germany","Greece","UK","US"))
ggplot(tempdata,aes(debtgdp,dRGDP,color=Country)) +
  geom_point()     # this produces the scatter plot

```

From here you can see that different countries appear to have quite different patterns.

Let's calculate average and median growth rates and debt to gdp ratios for these countries. To achieve this we will first group the data by Country `group_by(Country)` and then summarise the variables `dRGDP` and `debtgdp` (`summarise_at(c("dRGDP", "debtgdp"), ...)`) in the resulting groups by applying the mean and median function (`funs(mean, median, sd)`).

```{r}
tempdata %>% group_by(Country) %>% 
            summarise_at(c("dRGDP", "debtgdp"), funs(mean, median,sd)) %>% 
            print()
```

# Create categorical debt to GDP ratios - Reinhard-Rogoff analysis I

In order to replicate some of the analysis in Reinhard and Rogoff and the subsequent critique in Herndon et al. we will want to create categorical variables for the Debt to GDP ratio. In other words we want to 

```{r}
RRData <- RRData %>% mutate(dgcat = cut(RRData$debtgdp, breaks=c(0,30,60,90,Inf)))

RRData %>% group_by(dgcat) %>% 
            summarise_at("dRGDP", funs(mean, median)) %>% 
            print()
```

# Hypothesis Testing - means and differences in mean

Hypothesis testing is at the core of much empirical analysis. Here we will show how to perform hyothesis tests. We are using the `t.test` function. Often hypothesis tests will be testing hypotheses about the mean of a random variable. We also learn how to perform hypothesis on regression coefficients (often these are nothing else but ways to estimate means!).

Let's start by testing a hypothesis on the sample mean of the GDP growth rate (`dRGDP`) Let's say we want to test the hypothesis that the mean growth rate is 3.3% (note that 3.3% in the dat aiis represented as 3.3). Let's get one calculated and then diiscuss what we really did:

```{r}
t.test(RRData$dRGDP, mu=3.3)
```

You may remember that, in order to perform a hypothesis test you, the applied economist, has to set the null ($H_0$) and alternative ($H_A$) hypothesis. Here you set the null hypothesis that the mean (`mu`) is equal to 3 ($H_0$:`mu=3.3`). You also need to specify an alternative hypothesis. We didn't specify any so R (or better `t.test`) used it's default option, two sided alternative ($H_A: mu \neq 3.3$). 

The judgemment here is that it is possible that the true, yet unknown population mean is equal to 3.3. The test statistic is 1.4896 and the p-value is 0.1366. Only if the p-value is smaller than our chosen level of significance (often 0.01, 0.05 or 0.1) would we reject the null hypothesis $H_0$. 

Make sure you use the help function (type `?t.test` into the command window) or search for help (type "R t.test" into your favourite search engine) to understand more details of the function.

Let's calculate a one sided test with the alternative that ($H_A: mu > 3.3$):

```{r}
t.test(RRData$dRGDP, mu=3.3, alternative="greater")
```

Changing the alternative does not change the t-test but change the p-value. Now the p-value is 0.0683 and whether we reject or do not reject $H_0$ depends on our chosen significance level.

At this stage a we need to mention a huge caveat to the above tests. Our standard tests are based on the assumption that the data are identically and independently distributed. In some sense we have already seen that the this assumption cannot be defended with the data at hand. We saw the time-series plot above which showed that observations are correlated between years and ountries, so they are not independent observations. But let's, for the sake of this introduction, ignore this complication, but note that we need to interpret results with caution.

Above we sliced the data into different subsets. Let's use the debt level categories and compare means of growth between the subsets.

```{r}
temp_high <- RRData %>% filter(dgcat == "(90,Inf]")
temp_middle <- RRData %>% filter(dgcat == "(60,90]")

t.test(temp_high$dRGDP,temp_middle$dRGDP)
```

At the bottom of the output you can see the respective sample means (which replicates the sample means we printed earlier in a table). The hyothesis tested is $H_0: mu_{high}=mu_{middle}$ with the alternative that the respective population means are different ($H_a:mu_{high}\neq mu_{middle}$). As above we could have elected to test an alternative of $>$ or $<$ as well.

So yes there seems to be some negative correlation between the level of debt and the economic growth. So what was all the fuss about?

# Reinhard-Rogoff Analysis II

The reason why the work by Reinhard and Rogoff was so controversial was that it wasn't the numbers as reported above they published. They reported numbers after changing the dataset as follows:

1. They excluded early-postwar data for New Zealand, Australia and Canada, arguing that these data are atypical for later periods, essentially they are outliers

```{r}
## Selective treatment of early years

RRselective <- RRData %>% filter(!((Year<1950 & Country=="New Zealand") | (Year<1951 & Country=="Australia") | (Year<1951 & Country=="Canada") ))
```

2. A spreadsheet error resulted in data for the five countries (Australia, Austria, Belgium, Canada and Denmark) to not be included: 

```{r}
RRselective <- RRselective %>% filter(!( Country %in% c("Australia","Austria","Belgium","Canada","Denmark") ))
```

And now we repeat the mean and median calculations we performed above on teh complete dataset.

```{r}
RRselective %>% group_by(dgcat) %>% 
            summarise_at("dRGDP", funs(mean, median)) %>% 
            print()
```

This by itself changes the results somewhat (the mean growth rate of the highest indebted group drops from 2.17% to 1.69%) but the differences aren't all that dramatic. It is the combination of the above and the following third change that result in a dramatic change.

3. Observations are re-weighted according to the following scheme:

```{r}
RRselective2 <- RRselective %>%
                group_by(dgcat,Country) %>%                           # create category and country groups
                summarize( m1 = mean(dRGDP, na.rm = TRUE)) %>%        # calculate average growth in each category
                group_by(dgcat) %>%                                   # group again by category
                summarize(  n = n(), mean = mean(m1, na.rm = TRUE), median = median(m1, na.rm = TRUE)) %>%  # calculate mean in each category 
                print()
```

Clearly, this scheme delivers clearly lower average and median growth for the highest debt category.

The main difference in this new scheme is that not each country-year receives the same weight. The authors calculate groups of data by `dgcat` and `Country` and then calculate average growth rates. That means that for instance France receives one average growth rate for each of the the three lowest debt categories (not in any tear did it exceed the 90% threshold) and Germany receives an average growth for the two lowest debt categories (as it never exceeded the 60% threshold in the sample period). Then the country averages in each category are averaged again.

This may seem totally arbitrary

# R programming concepts used

* 