---
title: "Introduction to Handling Data"
subtitle: "ECON20222 - Lecture 1"
author: "Ralf Becker and Martyn Andrews"
date: "30 October 2018"
output: 
  beamer_presentation:
    includes:
      in_header: ../latex_template.tex  # use for full version
#     in_header: ../latex_student.tex  # use for student version
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## What is this course unit about?

\begin{itemize}
  \item Help you implement and interpret the main inference techniques used in Economics
  \item Focus on:
      \begin{itemize}
        \item causal inference
        \item the main pitfalls of time-series analysis
      \end{itemize}
\end{itemize}

## At the end of this unit ...

You will be able to:

\begin{itemize}
  \item Do intermediate data work in R
  \item Confidently apply regression analysis in R
  \item Apply more advanced causal inference techniques in R
  \item Find coding help for any new challenges in R
  \item Identify inference appropriate for the occasion
  \item Discuss strengths and weaknesses of particular empirical applications
  \item Interpret empirical results (with due caution!)
\end{itemize}

## What you need to do

To learn in this unit you need to:

\begin{figure}
	\centering
	\includegraphics[width=8cm]{Embrace.jpg}\\
\end{figure}

\begin{columns}
  \begin{column}{0.5\textwidth}
    \textcolor{student}{coding, cleaning data, struggling, amazement at what you can do}
  \end{column}
  \begin{column}{0.5\textwidth}
    \textcolor{student}{answering real questions, that there is not always a clear answer}
  \end{column}
    
\end{columns}


## Aim for today


\begin{columns}
  
  \begin{column}{0.5\textwidth}
    \underline{Statistics/Econometrics}
    \begin{itemize}
      \item Summary Statistics
      \item Difference between population and sample
      \item Hypothesis testing
      \item Graphical Data Representations
      \item Simple regression analysis
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    \underline{R Coding}
    \begin{itemize}
      \item Introduce you to R and RStudio
      \item How do I learn R
      \item Import data into R
      \item Perform some basic data manipulation
      \item Perform hypothesis tests
      \item Estimate a regression
    \end{itemize}
  \end{column}
    
\end{columns}


## Why Data Matter

\begin{columns}
  
  \begin{column}{0.5\textwidth}
    \begin{figure}
    	\centering
    	\includegraphics[width=3cm]{CarmenReinhart.jpg}\\
    \end{figure}

  \end{column}
  \begin{column}{0.5\textwidth}
    \begin{figure}
    	\centering
    	\includegraphics[width=3cm]{KennethRogoff.jpg}\\
    \end{figure}

  \end{column}
    
\end{columns}


\textcolor{student}
Average GDP growth rates (High Income countries - 1946 to 2009):

\begin{tabular}{|c|c|c|c|c|}
	\hline 
	Debts Category & 0,30] & (30,60] & (60,90] & (90,Inf] \\ 
	\hline 
	Avg Growth Rate &  \textcolor{student}{4.09\%} & \textcolor{student}{2.87\%}  &  \textcolor{student}{3.40\%} & \textcolor{student}{-0.02\%}  \\ 
	\hline 
\end{tabular} 

## Introduce R/R-Studio


## Write Code Files or the Basic Workflow

```{r, echo = FALSE, error = FALSE, message = FALSE, results = FALSE, warning = FALSE}
library(tidyverse)    # for almost all data handling tasks
library(readxl)       # to import Excel data
library(ggplot2)      # to produce nice graphiscs
library(stargazer)    # to produce nice results tables
RRData <- read_excel("RRdata.xlsx")
str(RRData)  # prints some basic info on variables
```

## The data

Upload the data
and check years and countries

## Scatter plot of the data