---
title: "Comp Lab 3"
author: "Ralf Becker"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This can be taken into account by adding "state fixed effcets".

```{r, eval = TRUE}
mod4 <- lm(Age.Adjusted.Rate~law.officers.pc+ur+alcc.pc+State,data=data2)
stargazer(mod2,mod3,mod4,type = "text", keep = c("law.officers.pc","ur","alcc.pc"))
```

Before we discuss what this regression actually did, we can observe something really interesting. 

* The $R^2$ of the regression has dramatically increased to 84%
* All three coefficients we previously estimated have now been estimated to be very statistically significant.
* The sign of the coefficient to `law.officers.pc` has changed, it is now negative.
* A lot of coefficients have been added to the model, one for each state, but for Alabama.

What the addition of the factor variable `State` does is that it includes a dummy variable for each state. For instance `StateArizona` is a variable that take sthe value 


As an aside, if you estimate a model that has heaps of coefficients, like `mod4` but you are mainly interested in a subset of these coefficients then you can ask `stargazer` to only show that particular subset. You can try the following command. 

```{r, eval = FALSE}
stargazer(mod2,mod3,mod4,type = "text", keep = c("law.officers.pc","ur","alcc.pc"))
```
