---
title: "Time-Series Modelling/Forecasting"
subtitle: "ECON20222 - TS Revision"
author: "Ralf Becker and Martyn Andrews"
date: "May 2021"
output: 
  beamer_presentation:
    includes:
#      in_header: ../latex_template.tex
      in_header: ../latex_student.tex  # use for student version
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```


## Time-Series Properties

We model economic time-series in order to:

\begin{itemize}
  \item understand the dynamics in \textbf{one} or between \textbf{several} time-series
  \item Forecast the \textbf{one or several} time-series 
  \item understand causal relationships between \textbf{several} time-series
\end{itemize}

```{r, echo = FALSE}
library(tidyverse)
library(ggplot2)
library(pdfetch)
library(xts)
library(AER)          # access to HS robust standard errors
library(stargazer)
source("stargazer_HC.r")  # includes the robust regression display
source("stargazer_HAC.r")  # includes the Newey-West standard errors
library(gridExtra)  # required for the combination of ggplots
```

## Import some data into R

Income tax and compulsory social contributions (NICs): £m CPNSA, monthly observations

```{r, echo = FALSE, results = 'hide'}
# Income tax and compulsory social contributions (NICs): £m CPNSA
tax <- pdfetch_ONS("KSS8","PUSF")  
periodicity(tax)   # check data availability
names(tax) <- "Income.Tax" # give a sensible name
```

```{r, echo = FALSE}
# we prepare the data for being kept in long format
# that is useful for plotting in ggplot
tax_l <- data.frame(index(tax),stack(as.data.frame(coredata(tax))))
# Give sensible names to columns
names(tax_l)[1] <- "Date"   # first col will have date
names(tax_l)[2] <- "Value"  # second col will have value
names(tax_l)[3] <- "id"     # third col will have series name
```

```{r, echo = FALSE}
ggplot(tax_l,aes(x =Date, y=Value)) + 
  geom_line(colour = "blue",size = 1.5) +
  ggtitle("UK income tax receipts") +
  theme_bw()
```


## The autocorrelation function (ACF)

The ACF expresses how observations are correlated to observations 1, 2, 3 or $k$ observations prior, $\rho_k$.


```{r, echo = FALSE, fig.height = 2.5, fig.width = 3.5}
temp_acf <- acf(tax)
```



## Stationary and Nonstationary Series

The ACF expresses how persistant a series is.

* A series that is extremely persistant is called a **nonstationary** series.  
* A series that is not very persistant is called a **stationary** series.

Here: The tax receipt series is nonstationary. 

* In general series with a time-trend are nonstationary
* **BUT** there is a huge grey area inbetween.  

Formal statistical tests exist (ot covered here). Here we eye-ball the series and look at how slowly the ACF converges to 0.

## Transformations

An important time-series transformation we consider is that of **differencing** a series (or log differencing)

```{r, fig.height = 2.5, fig.width = 3.5}
# we multiply by 100 to express in percentage points, i.e. 0.5 is 0.5% or 0.005
dtax <- 100*diff(log(tax), lag = 1)
temp_acf <- acf(dtax, na.action = na.pass)
```

## Transformations

Let's difference with ```lag = 12```, e.g. the difference between January and January receipts.

```{r, fig.height = 2.5, fig.width = 3.5}
# we multiply by 100 to express in percentage points, i.e. 0.5 is 0.5% or 0.005
dtax <- 100*diff(log(tax), lag = 12)
temp_acf <- acf(dtax, na.action = na.pass)
```

Differencing can turn nonstationary series into a stationary series.

## Spurious Regression

If we run a regression involving nonstationary variables...

Let's get some datasets fron EUROSTAT (using `pdfetch_EUROSTAT`).

* % of agricultural area Total fully converted and under conversion to organic farming in Germany [`Org`]  
* Thousands of passengers travelling to and from Norway by boat [`Pass`]  
* % of population with tertiary education in Italy [`Tert`]  
* Hospital Discharges, Alcoholic liver disease (in Thousands) in France [`Alc`] 


## Spurious Regression

All possible combinations of simple regressions between the four variables.

\scriptsize

\begin{table}
   \centering
\caption{Regression statistics} \label{Table_SpurReg}
\begin{tabular}{cccccc}
\hline \hline
&  & Exp. Var &  &  & \\
Dep.Var & & Org & Pass  & Tert & Alc \\ 
\hline 
Org & $\hat{\gamma}_1$ &  & 0.001 & 0.567*** & -0.125*** \\ 
& $se_{\hat{\gamma}_1}$ &  & (0.0005) & (0.031) & (0.009) \\ 
& $R^2$  &  & 0.127  & 0.954 & 0.928  \\ \hline
Pass & $\hat{\gamma}_1$ & 196.720 &  & 196.720 & -17.717 \\ 
& $se_{\hat{\gamma}_1}$ &  (137.969) &  & (137.969) & (18.298) \\ 
& $R^2$ & 0.127  &  & 0.188 & 0.067 \\ \hline
Tert & $\hat{\gamma}_1$ & 1.681*** & 0.001* &  & -0.215*** \\ 
& $se_{\hat{\gamma}_1}$ & (0.093) & (0.001) &  & (0.013) \\ 
& $R^2$ & 0.954 & 0.188 &  & 0.951 \\ \hline
Alc & $\hat{\gamma}_1$ & -7.421*** & -0.004 & -4.420*** &  \\ 
& $se_{\hat{\gamma}_1}$ & (0.532) & (0.004) & (0.259) &  \\ 
& $R^2$ & 0.928 & 0.067 &0.951  &  \\ \hline
\end{tabular} \\

Note:  *$p<0.1$; **$p<0.05$; ***$p<0.01$, \\
Newey-West standard errors in parenthesis
\end{table}

\normalsize


## Adding dynamic effects and Forecasting models

Note:   
* $\Delta ur$: change in unemployment rate  
* $\Delta rGDP$: change in real GDP

\begin{eqnarray*}
\Delta ur_t &=& \alpha_0 + \alpha_1 \Delta ur_{t-1} + \alpha_2 \Delta ur_{t-2} + ... + \alpha_p \Delta ur_{t-p} + \\
&& ~ \beta_1 ~ \Delta rGDP_{t-1} + ... +\beta_k ~ \Delta rGDP_{t-k} + u_t
\end{eqnarray*}

**No** contemporaneous terms on the right hand side! We can produce forecasts for period $t$ only having information at time (say) $t-1$.

These models can also be used to perform granger causality testing (Demo Class 5 and some Group Projects).

By taking out all the $\Delta rGDP$ we reduce the model to an **autoregressive model (AR)** which often produces very useful short-term forecasts.

## Forecasting Issues

* You may model a differenced series (e.g. $\Delta ur$) but still be interested in forecasting the level ($ur$)
* Often we forecast from AR models
* Order selection by information criterion
* If you have several forecasting models then they can be compared by comparing the RMSE (or other measures which we did not cover).
